
```yaml
---
title: "Lab 1: Reproducible Data Visualization Workflow"
author: "Chloe Smith"
format: 
html:
  toc: true
  code-copy: true
  code-tools: true
execute:
  warning: false
  message: false
---
```

# Part 3: Load the Required Packages

```{r}
#| label: load-packages


library("tidyverse")
library("here")
library("fs")
```

# Part 4: Confirm the Project Root

```{r}
#| label: confirm-project-root

here()
```

Answer:

1. What path was printed?
**"/Users/chlsmith-25/Documents/DataVisualization/Labs"**

2. Does it point to the correct course project?
**Yes, the path points to the correct location for the course project and the Labs sub-folder where this assignment is stored.**

3. Why is working from an R Project preferable to manually setting the working directory?
**Working from an R Project keeps project files organized under a consistent directory and makes file paths easier to manage. This improves reproducibility because files can be referenced relative to the project instead of requiring manual or sporadic changes to the directory** 

# Part 5: Create the Folder Structure

**The following labeled folders were created:**
**-Data**
**-Figures**
**-Scripts**
**-Tables**
**-Assignments**
**-Labs**
**-Portfolio**

## Added Tool for my Personal Understanding

**The required course folders were created previously in the main DataVisualization folder. Because here() now correctly identifies the main DataVisualization folder as the project root, each required directory can be referenced directly without moving up a level.**

```{r}
#| label: create-folder-structure

folders <- c(
  "data",
  "figures",
  "scripts",
  "tables",
  "assignments",
  "labs",
  "portfolio"
)
folders %>%
  walk(~ dir_create(here(.x)))
```

# Part 6: Inspect the Folder Structure

```{r}
#| label: inspect-folder-structure1

dir_ls()

```
## Adjusting the code to list top level project contents
```{r}
#| label: inspect-folder-structure2
dir_ls(here(".."))
```

# Part 7: Create a Simulated Public Health Dataset

Create a reproducible county-level dataset with the following variables:

- Include 12 fictional counties.
**Schnauzer County, Quarto County, Clifton County, Creek County, Midtown County, Cascade County, Mustard County, Hydrant County, Dove County, French County, Green County, Corinth County**

- Include three regions: North, Central, and South.
**North Region, Central Region, South Region**

# Because we don't want to manually generate random values for 12 counties, we're going to generate codes for random selection. Each requested category will have its own code chunk instead of one large chunk to reduce errors and gradually build the data-set.

## Counties
```{r}
#| label: create-simulated-health-data

set.seed(716)

county_health_data <- tibble(
  county = c(
    "Schauzer County",
    "Quarto County",
    "Clifton County",
    "Creek County",
    "Midtown County",
    "Cascade County",
    "Mustard County",
    "Hydrant County",
    "Dove County",
    "French County",
    "Green County",
    "Corinth County"
),
)
```

## Regions
```{r}
#| label: creating-regions


county_health_data <- county_health_data %>%
  mutate(
region = rep(
    c("North", "Central", "South"),
    each = 4
  )
)
```

## Populations
**Assigning each county a random adult population between 40,000 and 180,000**
```{r}
#| label: assinging-random-population-amounts-to-counties


county_health_data <- county_health_data %>%
  mutate(
adult_population = sample(
  40000:180000,
  12
  )
)
```

## Vaccination Rates
**Assigning vaccination coverage between 62% and 94%**
```{r}
#| label: Assigning-vaccination-coverage-between-fixed-percentages


county_health_data <- county_health_data %>%
  mutate(
vaccination_rate = round(
  runif(
    12,
    min = 62,
    max = 94
  ),
  1
)
)
```

## Asthma Rates Per 100,000
**Generating asthma visits proportionally with "runif"**
**Calculating asthma visit rate per 100,000 people**
```{r}
#| label: Calculating-asthma-rates-per-population-proportion


county_health_data <- county_health_data %>%
  mutate(
    asthma_visits = round(
      adult_population * runif(
        12,
        min = 0.004,
        max = 0.018
     )
      ),
    
  asthma_visit_rate = round(
    asthma_visits / adult_population * 100000,
    1
  )
  )
```

# Part 8: Save the Dataset to the Data folder

```{r}
#| label: save-simulated-health-data


write_csv(
  county_health_data,
  here("Data", "county_health_data.csv")
)
```

# Part 9: Removing and Re-Importing the Dataset

```{r}
#| label: remove-health-data

rm(county_health_data)
```

```{r}
#| label: import-health-data

county_health_data <- read_csv(
  here("Data", "county_health_data.csv")
)
```

# Part 10: Inspect the Data

```{r}
#| label: inspect-health-data

glimpse(county_health_data)
```


## Creating a one row summary
**Summary will contain number of counties, mean vaccination rate, and median asthma visit rate**

```{r}
#| label: summarize-health-data

county_health_data %>%
  summarize(
    number_of_counties = n(),
    mean_vaccination_rate = mean(vaccination_rate),
    median_asthma_visit_rate = median(asthma_visit_rate)
  )
```

1. What does one row represent?
**One row represents one county and the correlating data.**

2. How many variables are included?
**There are three variables included. in the summary. There are six variables total.**

3. Which variables are categorical?
**None of the variables shown are categorical. However, considering the entire data set, county name and region are categorical. **

4. Which variables are numeric?
**All three shown variables are numeric. Considering the entire data set, number of counties, mean vaccination rate, and median asthma visit rate are numeric**

5. Which variable should be used to compare asthma burden fairly across counties?
**Asthma visit rate should be used because it expresses asthma visits per 100,000 adults.**

# Part 11: Create a Ranked Vaccination Bar Chart

## Rerunning Package Chunk
```{r}
#| label: Reloading-datasets


library(tidyverse)
library(here)
library(fs)
```


## Creating and saving the graph under "Vaccination Plot"
```{r}
#| label: create-vaccination-plot
#| fig-cap: "Simulated vaccination coverage by county."
#| fig-alt: "Horizontal bar chart ranking fictional counties by simulated vaccination coverage."

vaccination_plot <- ggplot(
  county_health_data,
  aes(
    x = vaccination_rate,
    y = reorder(county, vaccination_rate)
  )
) +
  geom_col()
```

## Adding the 80% line
```{r}
#| label: Add-percentage-line


vaccination_plot <- vaccination_plot +
  geom_vline(
    xintercept = 80,
    linetype = "dashed"
  )
```

## Adding Labels
```{r}
#| label: Add-labels

vaccination_plot <- vaccination_plot +
    labs(
    title = "Vaccination Coverage Across Fictional Counties",
    subtitle = "County vaccination coverage compared with an 80% reference level",
    x = "Vaccination Coverage (%)",
    y = NULL,
    caption = "Values are simulated."
  ) +
  theme_minimal()

vaccination_plot
```

### Creating a target category
**Adding a variable to indicate whether each county is at/above 80% or below 80%**
```{r}
#| label: creating-percentage-variable


county_health_data <- county_health_data %>%
  mutate(
    vaccination_status = if_else(
      vaccination_rate >= 80,
      "Meets 80% Target",
      "Below 80% Target"
    )
  )
```

### Updating the graph
```{r}
#| label: updating-graph


vaccination_plot <- ggplot(
  county_health_data,
  aes(
    x = vaccination_rate,
    y = reorder(county, vaccination_rate),
    fill = vaccination_status
  )
) +
  geom_col()
```

### Color coding the graph
```{r}
#| label: Color-coding-graph
vaccination_plot <- vaccination_plot +
  scale_fill_manual(
    values = c(
      "Below 80% Target" = "#DEF67C",
      "Meets 80% Target" = "#E0F0FF"
    ),
 
  name = "Vaccination Status"
  )

vaccination_plot
```

### Adding black borders
```{r}
#| label: Adding-black-borders


vaccination_plot <- ggplot(
  county_health_data,
  aes(
    x = vaccination_rate,
    y = reorder(county, vaccination_rate),
    fill = vaccination_status
  )
) +
  geom_col(
    color = "black"
  ) +
  scale_fill_manual(
    name = "Vaccination Status",
    values = c(
      "Below 80% Target" = "#DEF67C",
      "Meets 80% Target" = "#E0F0FF"
    )
  )

vaccination_plot
```

### Changing the axes
```{r}
#| label: Changing-graph-axes


vaccination_plot <- ggplot(
  county_health_data,
  aes(
    x = vaccination_rate,
    y = reorder(county, vaccination_rate),
    fill = vaccination_status
  )
) +
  geom_col(
    color = "black"
  ) +
  scale_fill_manual(
    name = "Vaccination Status",
    values = c(
      "Below 80% Target" = "#DEF67C",
      "Meets 80% Target" = "#E0F0FF"
    )
  ) +
  labs(
    x = "Vaccination Coverage (%)",
    y = "County"
  )

vaccination_plot
```

### Re Adding the 80% reference line and subtitles
```{r}
#| label: readding-reference-line-and-subtitles


vaccination_plot <- vaccination_plot +
  geom_vline(
    xintercept = 80,
    linetype = "dashed"
  ) +
  labs(
    title = "Vaccination Coverage Across Fictional Counties",
    subtitle = "County vaccination coverage compared with an 80% reference level",
    caption = "Values are simulated."
  ) +
  theme_minimal()

vaccination_plot
```

# Part 12: Interpret the Figure

1. What pattern does the figure show?
2. Which counties appear furthest below the target?
3. What context would be required before using this graph for a real decision?

**The figure shows differences in simulated vaccination coverage across the 12 counties, with only Corinth County and French County reaching the 80% target. Green County and Dove County are furthest below the target, while Quarto County comes very close to meeting it. Before using this graph to guide a real decision, more information would be needed about the population of each county, the source and quality of the vaccination data, and factors that could contribute to differences in coverage.**

# Part 13: Save the Vaccination Figure

## Reloading the libraries 
```{r}
#| label: reloading-libraries


library(tidyverse)
library(here)
library(fs)
```

```{r}
#| label: save-vaccination-plot

ggsave(
  filename = here("..", "Figures", "lab01_vaccination_coverage.png"),
  plot = vaccination_plot,
  width = 8,
  height = 6,
  units = "in",
  dpi = 300
)
```


# Part 14: Independent Visualization

## Creating one additional visualization using the same dataset.

## Scatterplot of Adult Population

**This particular choice will let us form a visualization that is a different graph type and also shows a relationship.**
```{r}
#| label: independent-visualization
#| fig-cap: "Independent visualization created by the student."

independent_plot <- ggplot(
  county_health_data,
  aes(
    x=adult_population,
    y = asthma_visits
  )
) +
  geom_point(
    size = 4,
    color = "#5d3847"
  ) +
  geom_smooth(
    METHOD = "lm",
    se = TRUE,
    color = "#E30B5D"
  ) +
  labs(
    title = "Adult Population and Asthma Visits Across Fictional Counties",
    subtitle = "Relationship between county population size and simulated asthma-related visits",
    x = "Adult Population",
    y = "Asthma Visits",
    caption = "Data are simulated."
  ) +
  theme_minimal()

independent_plot
```

# Part 15: Save the Independent Figure

```{r}
#| label: save-independent-visualization
ggsave(
  filename = here("Figures", "lab01_independent_visualization.png"),
  plot = independent_plot,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300
)

```

# Part 16: Tidy Coding Check

**`%>%` is used instead of the native pipe, one major function appears per line, arguments are indented consistently, object names are descriptive, no `setwd()` appears, no absolute file paths appear and all code chunks have labels.**

# Part 17: Git and GitHub

**This portion of the lab required more troubleshooting than originally anticipated because I had not completed all of the recommended Git and GitHub preparation in advance. As a result, I encountered several issues involving project organization, relative file paths, repository initialization, file staging, and GitHub authentication. These challenges ultimately provided a practical demonstration of why establishing the project structure and version-control environment early is an important part of a reproducible workflow.**

**The project was initially associated with an incorrect project location, which caused here() to reference the wrong directory. After correcting the project location and relative paths, the R environment was restarted, required packages and objects were reloaded, the code was rerun, and the Quarto document was rendered again. Rendering also helped identify and correct issues such as duplicate chunk labels and a syntax error that may otherwise have remained unnoticed.**

**Git was then initialized for the correct project directory. Because the directory contained files unrelated to this lab, I reviewed the untracked files and selectively staged only the five files associated with Lab 1. These were committed to the main branch using the message Complete Lab 1 reproducible workflow.**

**Finally, I created and connected the local repository to a GitHub repository. Initial attempts to authenticate through HTTPS using a personal access token were unsuccessful, so I configured SSH authentication instead. An ED25519 SSH key was generated locally, added to GitHub, and the repository remote was changed from HTTPS to SSH. The project was then successfully pushed to GitHub and verified online.**

**Overall, this process reinforced that reproducibility involves more than writing code that runs. It also depends on consistent project organization, relative file paths, a clean computing environment, intentional version control, and properly configured tools for sharing the work. Completing the recommended setup earlier would have prevented several of these complications, but troubleshooting them helped me better understand how each component of the workflow functions together.**

# Reflection Questions


## Question 1

Why is reproducibility important in public health research and practice?

**Research findings and data may be used to inform programs, policies, and decisions that affect entire populations. A reproducible workflow allows others to understand how results were produced, verify the process, and repeat the analysis when necessary. It also helps ensure that the work can still be understood and recreated even after time has passed or the project has been transferred to someone else.**  

## Question 2

How does a consistent project structure improve collaboration and review?

**Consistent project structure makes it easier for collaborators and reviewers to understand where data, figures, scripts, tables, and other materials are stored without having to search through an entire project. It also creates a shared organizational system, which reduces confusion when multiple people are working with the same files. This lab showed me that organization is not just about preference; the location and structure of files can directly affect whether code runs correctly.**  

## Question 3

Why are relative paths preferable to absolute paths?

**They reference files based on their location within the project rather than a location that is specific to one person's computer. An absolute path may work on my computer but fail when another person opens the project because their username, folders, or file locations are different. Using relative paths makes the project more portable and allows the same code to work across different computers as long as the project structure remains consistent.**  

## Question 4

How does GitHub contribute to reproducibility?

**GitHub contributes to reproducibility by providing a central location where project files and changes can be stored and tracked over time. Git commits create a history of the work, which makes it possible to see what changed and when instead of relying on multiple disconnected versions of the same file. GitHub also allows collaborators or reviewers to access the same materials used to produce an analysis, which supports transparency and makes the work easier to reproduce.**  

## Question 5

What was the most challenging part of the lab?

**The most challenging part of the lab was configuring the project and Git/GitHub workflow, particularly because I had not completed all of the recommended setup in advance. I encountered issues involving the project location, relative paths, an empty R environment after restarting, selective file staging, and GitHub authentication. Although troubleshooting these issues took additional time, it helped me understand why the setup steps matter and how the different parts of a reproducible workflow depend on one another. By the end of the lab, I was able to rerun the analysis from a clean environment, render the document successfully, selectively commit the required files, and push the completed project to GitHub using SSH authentication.**  

# Deliverables

Submit or push:

- [ ] completed student lab QMD
- [ ] rendered HTML file
- [ ] `data/county_health_data.csv`
- [ ] `figures/lab01_vaccination_coverage.png`
- [ ] `figures/lab01_independent_visualization.png`
- [ ] completed reflections
- [ ] meaningful Git commit
- [ ] updated GitHub repository

# Submission Checklist

Before submitting, confirm:

- [ ] the project opens from the `.Rproj` file
- [ ] the QMD renders without errors
- [ ] all required folders exist
- [ ] the dataset is inside `data/`
- [ ] relative paths are used
- [ ] both visualizations display correctly
- [ ] both figures are saved inside `figures/`
- [ ] reflections are complete
- [ ] work is committed and pushed
- [ ] no private or sensitive information was uploaded


